Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
edits/comment in keynote section

WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
Department of Astronomy, University of Illinois titled \emph{Why Sustain
Scientific Software?}. Turk is a prolific practitioner of scientific software
and has extensive experiences working on large collaborative projects employing
modern computing tools. He is also a co-organizer and champion of all the WSSSPE
events so far.

In his keynote address, Turk recapped the course of development of WSSSPE
workshops alongside with his career from a postdoc to an academic over the past
few years. The first WSSSPE workshop was at the Supercomputing conference (SC13)
in 2013, but he observed that the notion of sustainable scientific software drew
in an audience beyond supercomputing. In the following year, WSSSPE1.1 at SciPy
had speakers talking about how software has been sustained inside the scientific
Python community. WSSSPE2 at SC14 had breakout group discussions coming up with
actionable items, and WSSSPE2.1 at SciPy 2015 was similar. Turk noted the
different atmosphere of the surrounding large conferences, despite similar
WSSSPE participants.

WSSSPE3 broke free of the traditional Supercomputing Conference environment this
year, and in Turk's words, this spoke to the fact that scientific software comes
from many different types of inquiries, deployment, strategies for maintenance,
users; and ways of measuring the value of a piece of software. It appeared to
Turk that the supercomputing conferences generally adopt some top-down
approaches, whereas SciPy more often than not uses more bottom-up systems. The
essential messages perceived were also often bipolar: software getting harder in
supercomputing communities with exascale computing and optimization issues in
mind, but software becoming better in the SciPy group with emerging tools such
as Jupiter and productivity packages for research workflow. Admitting such
comparisons are somewhat unfair generalizations, Turk reminded the audience that
the different approaches bring different types of ideas to the table, but he
welcomed WSSSPE3 being conducted outside existing preconceptions.

@@

Turk then moved onto various different definitions of sustainability. His
favorite one was keeping up with bug reports, in which case even if no new
features were added, the software remains sustainable. Another definition of
sustainability Turk mentioned was adding of new features, or maintaining the
software for a long period of time such as the cases of TeX or LaTeX with
community help. A notion Turk heard often at supercomputing conferences was that
sustainable software continued to work on new architectures. Yet another metric
was people continuing to be able to learn how to use and apply the software. A
funder Turk heard talked about sustainability as continuing to get funded. Turk
also recalled that Greg Wilson, among others, said in WSSSPE1.1 that his view of
sustainable software was software that continued to give the same results over
time. A last measure of sustainability Turk presented was the ability to
transition between different people developing and using a piece of software.

At WSSSPE1, several models were presented for ensuring sustainability. Turk
considered that a familiar one was a funded piece of software where an external
agency provided funds to a group that was not necessarily exclusively working on
and developing a piece of software, keeps it going, and provides that out to the
scientific community. The model of productized software, in which a piece of
software has grown into something that individuals rely on to the point that
research groups or people are willing to support it at some pay rate, for
instance, subscription to use cloud services that deploy a piece of software, or
purchase of a piece of software. A final model Turk felt conflicted about is a
volunteer model that is traditional old-school---not modern-day open
source---developement.

Turk discussed whether productizing scientific software was synonoymous with
being sustsainable and self-sufficient. He thought it was not necessarily the
case and furthermore, it could lead to a divergence of interest between users
and developers.

The volunteer model meant unpaid labor. On this note, Turk recommended Ashe
Dryden's blog
post\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}
on the ethics of unpaid labor and the open source software community, which
mentions GitHub as one's resume, for example. Often times, a person funded to
work on some scientific project full time has a small amount of time for working
on a piece of software necessary for the project. However, a researcher's ability
to participate in that volunteer community is not always the same and it is not
always aligned with the research projects. From Turk's experience, we cannot
always rely on unpaid labor and volunteer time to sustain a piece of
software---that came down to the notions of the top-down and the bottom-up
approaches, the funded versus the grassroot.

>>>>

WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
Department of Astronomy, University of Illinois titled \emph{Why Sustain
Scientific Software?}. Turk is a prolific scientific software practitioner
and has extensive experiences working on large collaborative projects employing
modern computing tools. He also co-organizes and champions WSSSPE
events.

In his keynote address, Turk recapped the course of development of WSSSPE
workshops over the past
few years, alongside his career development from a postdoc to an academic. The first WSSSPE workshop was at the Supercomputing conference (SC13)
in 2013, but he observed that the notion of sustainable scientific software drew
in an audience beyond supercomputing. In the following year, WSSSPE1.1 at SciPy
had speakers talking about how software has been sustained inside the scientific
Python community. WSSSPE2 at SC14 had breakout group discussions coming up with
actionable items, and WSSSPE2.1 at SciPy 2015 was similar. Turk noted the
different atmosphere of the surrounding large conferences, despite similar
WSSSPE participants.

WSSSPE3 left the traditional Supercomputing Conference environment this
year, and in Turk's words, this spoke to the fact that scientific software comes
from many different types of inquiries, deployment, strategies for maintenance,
users, and ways of measuring the value of a piece of software. It appeared to
Turk that the supercomputing community generally adopt some top-down
approaches, whereas the SciPy community more often than not uses more bottom-up systems. The
essential messages perceived were also often bipolar: the supercomputing community thinks software is getting harder, with exascale computing and optimization issues in
mind, but the SciPy community thinks software becoming better, with emerging tools such
as Jupiter and productivity packages for research workflow. Admitting such
comparisons are somewhat unfair generalizations, Turk reminded the audience that
the different approaches bring different types of ideas to the table, and he
welcomed WSSSPE3 being conducted outside existing preconceptions.

@@

Turk then moved onto various different definitions of sustainability. His
favorite one was `keeping up with bug reports,' where even if no new
features were added, the software remains sustainable. Another definition of
sustainability Turk mentioned was `adding of new features', or `maintaining the
software for a long period of time' such as the cases of TeX or LaTeX with
community help. A notion Turk heard often at supercomputing conferences was that
sustainable software `continues to work on new architectures'. Yet another metric
was `people continuing to be able to learn how to use and apply the software'. A
funder Turk heard talked about sustainability as `continuing to get funded'. Turk
also recalled that Greg Wilson, among others, said in WSSSPE1.1 that his view of
sustainable software was software that `continued to give the same results over
time'. A last measure of sustainability Turk presented was `the ability to
transition between different people developing and using a piece of software.'

At WSSSPE1, several models were presented for ensuring sustainability. Turk
considered that a familiar one was a funded piece of software where an external
agency provided funds to a group who are not necessarily exclusively working on
and developing a piece of software, keeps it going, and provides it to the
scientific community. The model of productized software, in which a piece of
software has grown into something that individuals rely on to the point that
research groups or people are willing to support it with some amount of funding, for
instance, a subscription to use cloud services that deploy a piece of software, or
purchase of a piece of software. A final model Turk felt conflicted about is a
volunteer model that is traditional old-school---not modern-day open
source---development.

Turk discussed whether productizing scientific software was synonymous with
being sustainable and self-sufficient. He thought it was not necessarily the
case and furthermore, it could lead to a divergence of interest between users
and developers.

Turk reminded the audience that the volunteer model means unpaid labor. On this note, he recommended Ashe
Dryden's blog
post\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}
on the ethics of unpaid labor and the open source software community, which
mentions GitHub as one's resume, for example. \katznote{I don't think I understand the point in the previous sentence - maybe there are multiple points and this should be broken in multiple sentences?} Often times, a person funded to
work full time on some scientific project can spend a small amount of time for working
on a piece of software necessary for the project. However, all researchers' ability
to participate in that volunteer community are not always the same and may not
always be aligned with the research projects. From Turk's experience, we cannot
always rely on unpaid labor and volunteer time to sustain a piece of
software---this came down to the notions of the top-down and the bottom-up
approaches, the funded versus the grassroots.
